Counterfactual Training: Teaching Models Plausible and Actionable Explanations

### 2025-02-04 paper/body_ecml.tex
progress on methodology

\begin{abstract}
Counterfactual Explanations have emerged as a popular tool to explain
predictions made by opaque machine learning models: they explain how
factual inputs need to change in order for some fitted model to produce
some desired output. Much existing research has focused on identifying
explanations that are not only valid but also deemed desirable with
respect to the underlying data and stakeholder requirements. Recent work
has shown that under this premise, the task of learning desirable
explanations is effectively reassigned from the model itself to the
(post-hoc) counterfactual explainer. Building on that work, we propose a
novel model objective that leverages counterfactuals during the training
phase (ad-hoc) in order to minimize the divergence between learned
representations and desirable explanations. Through extensive
experiments, we demonstrate that our proposed methodology facilitates
training models that inherently deliver desirable explanations while
maintaining high predictive performance.
\keywords{Counterfactual Explanations \and Explainable
AI \and Representation Learning}
\end{abstract}

@@

Today's prominence of artificial intelligence (AI) has largely been
driven by advances in \textbf{representation learning}: instead of
relying on features and rules that are carefully hand-crafted by humans,
modern AIs are tasked with learning these representations from scratch,
guided by narrow objectives such as predictive accuracy
\cite{goodfellow2016deep}. Modern advances in computing have made it
possible to provide such AIs with ever greater degrees of freedom to
achieve that task, which has often led them to outperform traditionally
more parsimonious models. Unfortunately, in doing so they also learn
increasingly complex and highly sensitive representations that we can no
longer easily interpret.

@@

Part of the problem is that high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives \cite{wilson2020case}. Based purely on predictive
performance, these solutions may seem to provide compelling explanations
for the data, when in fact they are based on purely associative,
semantically meaningless patterns. This poses two related challenges:
firstly, it makes these models inherently opaque, since humans cannot
simply interpret what type of explanation the complex learned
representations correspond to; secondly, even if we could resolve the
first challenge, it is not obvious how to mitigate models from learning
representations that correspond to meaningless and undesirable
explanations.

@@

To our surprise, the second challenge has not yet attracted any
consolidated research effort. Specifically, there has been no concerted
effort towards improving model \textbf{explainability}, which we define
here as the degree to which learned representations correspond to
explanations that are interpretable and deemed desirable by humans.
Instead, the choice has typically been to improve the capacity of XAI
tools to identify the subset explanations that are both desirable and
valid for any given model, independent of whether the learned
representations are also compatible with undesirable explanations
\cite{altmeyer2024faithful}. Fortunately, recent findings indicate that
explainability can arise as byproduct of regularization techniques aimed
at other objectives such as robustness, generalization and generative
capacity \cite{schut2021generating}.

Building on these findings, we introduce \textbf{counterfactual
training}: a novel regularization technique geared explicitly towards
aligning model representations with desirable explanations. Our
contributions are as follows:

@@

To the best of our knowledge, our proposed framework for counterfactual
training represents the first attempt to use counterfactual explanations
during training to improve model explainability. In high-level terms, we
define model explainability as the extent to which valid explanations
derived for an opaque model are also deemed desirable with respect to
the underlying data and stakeholder requirements. To make this more
concrete, we follow \cite{augustin2020adversarial} in tieing the concept
of explainability to the quality of counterfactual explanations that we
can generate for a given model. The authors show that counterfactual
explanations---understood here as minimal input perturbations that yield
some desired model prediction---are generally more meaningful if the
underlying model is more robust to adversarial examples. We can make
intuitive sense of this finding when looking at adversarial training
(AT) through the lens of representation learning with high degrees of
freedom: by inducing models to ``unlearn'' representations that are
susceptible to worst-case counterfactuals (i.e.~adversarial examples),
AT effectively removes some undesirable explanations from the solution
space.

@@

Improving the adversarial robustness of models is not the only path
towards aligning representations with desirable explanations. In a work
closely related to this one, \cite{altmeyer2024faithful} show that
explainability can be improved through model averaging and refined model
objectives. The authors propose a way to generate counterfactuals that
are maximally \textbf{faithful} to the model in that they are consistent
with what the model has learned about the underlying data. Formally,
they rely on tools from energy-based modelling to minimize the
divergence between the distribution of counterfactuals and the
conditional posterior over inputs learned by the model. Their proposed
counterfactual explainer, \emph{ECCCo}, yields desirable (or
\textbf{plausible}) explanations if and only if the underlying model has
learned representations that align with them. They find that both deep
ensembles \cite{lakshminarayanan2016simple} and joint energy-based
models (JEMs) \cite{grathwohl2020your} tend to do well in this regard.

Once again it helps to look at these findings through the lens of
representation learning with high degrees of freedom. Deep ensembles are
approximate Bayesian model averages, which are most called for when
models are underspecified by the available data \cite{wilson2020case}.
Averaging across solutions mitigates the aforementioned risk of relying
on a single locally optimal representations that corresponds to
semantically meaningless explanations for the data. Previous work by
\cite{schut2021generating} similarly found that generating desirable
(``interpretable'') counterfactual explanations is almost trivial for
deep ensembles that have also undergone adversarial training. The case
for JEMs is even clearer: they involve a hybrid objective that induces
both high predictive performance and generative capacity
\cite{grathwohl2020your}. This is closely related to the idea of
aligning models with desirable explanations and has inspired our
proposed counterfactual training objective, as we explain in
Section~\ref{sec-method}.

@@

@@

~

@@

>>>>

\begin{abstract}
Counterfactual Explanations have emerged as a popular tool to explain
predictions made by opaque machine learning models: they explain how
factual inputs need to change in order for some fitted model to produce
some desired output. Much existing research has focused on identifying
explanations that are not only valid but also deemed plausible and
desirable with respect to the underlying data and stakeholder
requirements. Recent work has shown that under this premise, the task of
learning plausible explanations is effectively reassigned from the model
itself to the (post-hoc) counterfactual explainer. Building on that
work, we propose a novel model objective that leverages counterfactuals
during the training phase (ad-hoc) in order to minimize the divergence
between learned representations and plausible explanations. Through
extensive experiments, we demonstrate that our proposed methodology
facilitates training models that inherently deliver plausible
explanations while maintaining high predictive performance.
\keywords{Counterfactual Explanations \and Explainable
AI \and Representation Learning}
\end{abstract}

@@

Today's prominence of artificial intelligence (AI) has largely been
driven by advances in \textbf{representation learning}: instead of
relying on features and rules that are carefully hand-crafted by humans,
modern machine learning (ML) models are tasked with learning these
representations from scratch, guided by narrow objectives such as
predictive accuracy \cite{goodfellow2016deep}. Modern advances in
computing have made it possible to provide such models with ever greater
degrees of freedom to achieve that task, which has often led them to
outperform traditionally more parsimonious models. Unfortunately, in
doing so they also learn increasingly complex and highly sensitive
representations that we can no longer easily interpret.

@@

Part of the problem is that high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives \cite{wilson2020case}\footnote{For clarity: we follow
  standard ML convention in using ``degrees of freedom'' to refer to the
  number of parameters estimated from data.}. Based purely on predictive
performance, these solutions may seem to provide compelling explanations
for the data, when in fact they are based on purely associative,
semantically meaningless patterns. This poses two related challenges:
firstly, it makes these models inherently opaque, since humans cannot
simply interpret what type of explanation the complex learned
representations correspond to; secondly, even if we could resolve the
first challenge, it is not obvious how to mitigate models from learning
representations that correspond to meaningless and implausible
explanations.

@@

To our surprise, the second challenge has not yet attracted any
consolidated research effort. Specifically, there has been no concerted
effort towards improving model \textbf{explainability}, which we define
here as the degree to which learned representations correspond to
explanations that are interpretable and deemed \textbf{plausible} by
humans (see Definition~\ref{def-explainability}). Instead, the choice
has typically been to improve the capacity of XAI tools to identify the
subset explanations that are both plausible and valid for any given
model, independent of whether the learned representations are also
compatible with implausible explanations \cite{altmeyer2024faithful}.
Fortunately, recent findings indicate that explainability can arise as
byproduct of regularization techniques aimed at other objectives such as
robustness, generalization and generative capacity
\cite{schut2021generating}.

Building on these findings, we introduce \textbf{counterfactual
training}: a novel regularization technique geared explicitly towards
aligning model representations with plausible explanations. Our
contributions are as follows:

@@

To the best of our knowledge, our proposed framework for counterfactual
training represents the first attempt to use counterfactual explanations
during training to improve model explainability. In high-level terms, we
define model explainability as the extent to which valid explanations
derived for an opaque model are also deemed plausible with respect to
the underlying data and stakeholder requirements. To make this more
concrete, we follow \cite{augustin2020adversarial} in tieing the concept
of explainability to the quality of counterfactual explanations that we
can generate for a given model. The authors show that counterfactual
explanations---understood here as minimal input perturbations that yield
some desired model prediction---are generally more meaningful if the
underlying model is more robust to adversarial examples. We can make
intuitive sense of this finding when looking at adversarial training
(AT) through the lens of representation learning with high degrees of
freedom: by inducing models to ``unlearn'' representations that are
susceptible to worst-case counterfactuals (i.e.~adversarial examples),
AT effectively removes some implausible explanations from the solution
space.

@@

Improving the adversarial robustness of models is not the only path
towards aligning representations with plausible explanations. In a work
closely related to this one, \cite{altmeyer2024faithful} show that
explainability can be improved through model averaging and refined model
objectives. The authors propose a way to generate counterfactuals that
are maximally \textbf{faithful} to the model in that they are consistent
with what the model has learned about the underlying data. Formally,
they rely on tools from energy-based modelling to minimize the
divergence between the distribution of counterfactuals and the
conditional posterior over inputs learned by the model. Their proposed
counterfactual explainer, \emph{ECCCo}, yields plausible explanations if
and only if the underlying model has learned representations that align
with them. They find that both deep ensembles
\cite{lakshminarayanan2016simple} and joint energy-based models (JEMs)
\cite{grathwohl2020your} tend to do well in this regard.

Once again it helps to look at these findings through the lens of
representation learning with high degrees of freedom. Deep ensembles are
approximate Bayesian model averages, which are most called for when
models are underspecified by the available data \cite{wilson2020case}.
Averaging across solutions mitigates the aforementioned risk of relying
on a single locally optimal representations that corresponds to
semantically meaningless explanations for the data. Previous work by
\cite{schut2021generating} similarly found that generating plausible
(``interpretable'') counterfactual explanations is almost trivial for
deep ensembles that have also undergone adversarial training. The case
for JEMs is even clearer: they involve a hybrid objective that induces
both high predictive performance and generative capacity
\cite{grathwohl2020your}. This is closely related to the idea of
aligning models with plausible explanations and has inspired our
proposed counterfactual training objective, as we explain in
Section~\ref{sec-method}.

@@

Counterfactual training combines ideas from adversarial training,
energy-based modelling and counterfactuals explanations with the
explicit objective of aligning representations with plausible
explanations that comply with user requirements. In the context of CE,
plausibility has broadly been defined as the degree to which
counterfactuals comply with the underlying data generating process
\cite{poyiadzi2020face, guidotti2022counterfactual, altmeyer2024faithful}.
Plausibility is a necessary but insufficient condition for using CE to
provide algorithmic recourse (AR) to individuals affected by opaque
models in practice. This is because for recourse recommendations to be
\textbf{actionable}, they need to not only result in plausible
counterfactuals but also be attainable. A plausible CE for a rejected
20-year-old loan applicant, for example, might reveal that their
application would have been accepted, if only they were 20 years older.
Ignoring all other features, this complies with the definition of
plausibility if 40-year-old individuals are in fact more credit-worthy
on average than young adults. But of course this CE does not qualify for
providing actionable recourse to the applicant. For our intents and
purposes, counterfactual training aims at improving model explainability
by aligning models with counterfactuals that meet both desiderata,
plausibility and actionability. Formally, we define explainability as
follows:

@@

Let \(M_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denote a supervised
classification model that maps from intuts to representations
\(\phi(x;theta)\) and finally to outputs. Assume that for any given
input-output pair \(\{\mathbf{x},\mathbf{y}\}\) there exists a
counterfactual
\(\mathbf{x}^{\prime} = \mathbf{x} + \Delta: M_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = M_\theta(\mathbf{x})\)
where \(\mathbf{y}^{+}\) denotes some target output. We say that
\(M_\theta\) is \textbf{explainable} to the extent that faithfully
generated counterfactuals are:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Plausible: \(\mathbf{x}^{\prime} \sim_{p} \mathcal{X}|\mathbf{y}^{+}\)
  with \(p \rightarrow 1\) as defined in \cite{altmeyer2024faithful}.
\item
  Actionable: permutations \(\Delta\) are subject to actionability
  constraints.
\end{enumerate}

@@

Actionability constraints in Definition~\ref{def-explainability} vary
and depend on the context in which \(M_\theta\) is deployed. In this
work, we focus on domain and mutability constraints for individual
features. We also limit ourselves to classification tasks in this work,
a limitation that we discuss in Section~\ref{sec-discussion}.

\subsection{Our Proposed Objective}\label{our-proposed-objective}

To train models with high explainability as defined in
Definition~\ref{def-explainability}, we propose the following objective,

\begin{equation}\phantomsection\label{eq-obj}{
\text{yloss}(M_\theta(x),y) + \lambda_{\text{div}} \text{div}({x^\prime},x;\theta) + \lambda_{\text{adv}} \text{advloss}(M_\theta({x^\prime}),y)
}\end{equation}

where \(\text{yloss}(\cdot)\) denotes any conventional classification
loss function (e.g.~crossentropy) that induces discriminative
(predictive) performance. The two additional components in
Equation~\ref{eq-obj} are explained in more detail below. For now they
can be sufficiently described as inducing explainability directly and
indirectly by penalizing: 1) the contrastive divergence,
\(\text{div}(\cdot)\), between counterfactuals \(x^\prime\) and observed
samples \(x\) and, 2) the adversarial loss, \(\text{advloss}(.)\), with
respect to counterfactuals. The tradeoff between the different
components can be governed by adjusting the strengths of the penalties
\(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\).

\subsubsection{Directly Inducing Explainability through Contrastive
Divergence}\label{directly-inducing-explainability-through-contrastive-divergence}

\subsubsection{Indirectly Inducing Explainability through Adversarial
Robustness}\label{indirectly-inducing-explainability-through-adversarial-robustness}

A reasonable choice for the latter is to define
\(\text{advloss}(M_\theta({x^\prime}),y;\varepsilon):=\text{yloss}(M_\theta({x^\prime}),y)\)

\subsection{Encoding Domain Knowledge}\label{encoding-domain-knowledge}

### 2025-02-04 paper/preprint/paper.tex
progress on methodology

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Parameters}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Counterfactual Training (A Preprint) }
\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Patrick
Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
Engineering, Mathematics and Computer Science\\Delft University of
Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Arie
van Deursen}\\Faculty of Electrical Engineering, Mathematics and
Computer Science\\Delft University of Technology\\\\\asep\textbf{Cynthia
C. S. Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\}
\date{}
\begin{document}
\maketitle
\begin{abstract}
Counterfactual Explanations have emerged as a popular tool to explain
predictions made by opaque machine learning models: they explain how
factual inputs need to change in order for some fitted model to produce
some desired output. Much existing research has focused on identifying
explanations that are not only valid but also deemed desirable with
respect to the underlying data and stakeholder requirements. Recent work
has shown that under this premise, the task of learning desirable
explanations is effectively reassigned from the model itself to the
(post-hoc) counterfactual explainer. Building on that work, we propose a
novel model objective that leverages counterfactuals during the training
phase (ad-hoc) in order to minimize the divergence between learned
representations and desirable explanations. Through extensive
experiments, we demonstrate that our proposed methodology facilitates
training models that inherently deliver desirable explanations while
maintaining high predictive performance.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Explanations \sep Explainable AI \sep 
Representation Learning

@@

Today's prominence of artificial intelligence (AI) has largely been
driven by advances in \textbf{representation learning}: instead of
relying on features and rules that are carefully hand-crafted by humans,
modern AIs are tasked with learning these representations from scratch,
guided by narrow objectives such as predictive accuracy
(\citeproc{ref-goodfellow2016deep}{I. Goodfellow, Bengio, and Courville
2016}). Modern advances in computing have made it possible to provide
such AIs with ever greater degrees of freedom to achieve that task,
which has often led them to outperform traditionally more parsimonious
models. Unfortunately, in doing so they also learn increasingly complex
and highly sensitive representations that we can no longer easily
interpret.

@@

Part of the problem is that high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives (\citeproc{ref-wilson2020case}{Wilson 2020}). Based purely on
predictive performance, these solutions may seem to provide compelling
explanations for the data, when in fact they are based on purely
associative, semantically meaningless patterns. This poses two related
challenges: firstly, it makes these models inherently opaque, since
humans cannot simply interpret what type of explanation the complex
learned representations correspond to; secondly, even if we could
resolve the first challenge, it is not obvious how to mitigate models
from learning representations that correspond to meaningless and
undesirable explanations.

@@

To our surprise, the second challenge has not yet attracted any
consolidated research effort. Specifically, there has been no concerted
effort towards improving model \textbf{explainability}, which we define
here as the degree to which learned representations correspond to
explanations that are interpretable and deemed desirable by humans.
Instead, the choice has typically been to improve the capacity of XAI
tools to identify the subset explanations that are both desirable and
valid for any given model, independent of whether the learned
representations are also compatible with undesirable explanations
(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
Fortunately, recent findings indicate that explainability can arise as
byproduct of regularization techniques aimed at other objectives such as
robustness, generalization and generative capacity Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}).

Building on these findings, we introduce \textbf{counterfactual
training}: a novel regularization technique geared explicitly towards
aligning model representations with desirable explanations. Our
contributions are as follows:

@@

To the best of our knowledge, our proposed framework for counterfactual
training represents the first attempt to use counterfactual explanations
during training to improve model explainability. In high-level terms, we
define model explainability as the extent to which valid explanations
derived for an opaque model are also deemed desirable with respect to
the underlying data and stakeholder requirements. To make this more
concrete, we follow Augustin, Meinke, and Hein
(\citeproc{ref-augustin2020adversarial}{2020}) in tieing the concept of
explainability to the quality of counterfactual explanations that we can
generate for a given model. The authors show that counterfactual
explanations---understood here as minimal input perturbations that yield
some desired model prediction---are generally more meaningful if the
underlying model is more robust to adversarial examples. We can make
intuitive sense of this finding when looking at adversarial training
(AT) through the lens of representation learning with high degrees of
freedom: by inducing models to ``unlearn'' representations that are
susceptible to worst-case counterfactuals (i.e.~adversarial examples),
AT effectively removes some undesirable explanations from the solution
space.

@@

Improving the adversarial robustness of models is not the only path
towards aligning representations with desirable explanations. In a work
closely related to this one, Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}) show that explainability can
be improved through model averaging and refined model objectives. The
authors propose a way to generate counterfactuals that are maximally
\textbf{faithful} to the model in that they are consistent with what the
model has learned about the underlying data. Formally, they rely on
tools from energy-based modelling to minimize the divergence between the
distribution of counterfactuals and the conditional posterior over
inputs learned by the model. Their proposed counterfactual explainer,
\emph{ECCCo}, yields desirable (or \textbf{plausible}) explanations if
and only if the underlying model has learned representations that align
with them. They find that both deep ensembles
(\citeproc{ref-lakshminarayanan2016simple}{Lakshminarayanan, Pritzel,
and Blundell 2017}) and joint energy-based models (JEMs)
(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) tend to do
well in this regard.

Once again it helps to look at these findings through the lens of
representation learning with high degrees of freedom. Deep ensembles are
approximate Bayesian model averages, which are most called for when
models are underspecified by the available data
(\citeproc{ref-wilson2020case}{Wilson 2020}). Averaging across solutions
mitigates the aforementioned risk of relying on a single locally optimal
representations that corresponds to semantically meaningless
explanations for the data. Previous work by Schut et al.
(\citeproc{ref-schut2021generating}{2021}) similarly found that
generating desirable (``interpretable'') counterfactual explanations is
almost trivial for deep ensembles that have also undergone adversarial
training. The case for JEMs is even clearer: they involve a hybrid
objective that induces both high predictive performance and generative
capacity (\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}). This
is closely related to the idea of aligning models with desirable
explanations and has inspired our proposed counterfactual training
objective, as we explain in Section~\ref{sec-method}.

@@

@@

@@

@@

>>>>

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{exercise}{Parameters}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Counterfactual Training (A Preprint) }
\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Patrick
Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
Engineering, Mathematics and Computer Science\\Delft University of
Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Arie
van Deursen}\\Faculty of Electrical Engineering, Mathematics and
Computer Science\\Delft University of Technology\\\\\asep\textbf{Cynthia
C. S. Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\}
\date{}
\begin{document}
\maketitle
\begin{abstract}
Counterfactual Explanations have emerged as a popular tool to explain
predictions made by opaque machine learning models: they explain how
factual inputs need to change in order for some fitted model to produce
some desired output. Much existing research has focused on identifying
explanations that are not only valid but also deemed plausible and
desirable with respect to the underlying data and stakeholder
requirements. Recent work has shown that under this premise, the task of
learning plausible explanations is effectively reassigned from the model
itself to the (post-hoc) counterfactual explainer. Building on that
work, we propose a novel model objective that leverages counterfactuals
during the training phase (ad-hoc) in order to minimize the divergence
between learned representations and plausible explanations. Through
extensive experiments, we demonstrate that our proposed methodology
facilitates training models that inherently deliver plausible
explanations while maintaining high predictive performance.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Explanations \sep Explainable AI \sep 
Representation Learning

@@

Today's prominence of artificial intelligence (AI) has largely been
driven by advances in \textbf{representation learning}: instead of
relying on features and rules that are carefully hand-crafted by humans,
modern machine learning (ML) models are tasked with learning these
representations from scratch, guided by narrow objectives such as
predictive accuracy (\citeproc{ref-goodfellow2016deep}{I. Goodfellow,
Bengio, and Courville 2016}). Modern advances in computing have made it
possible to provide such models with ever greater degrees of freedom to
achieve that task, which has often led them to outperform traditionally
more parsimonious models. Unfortunately, in doing so they also learn
increasingly complex and highly sensitive representations that we can no
longer easily interpret.

@@

Part of the problem is that high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives (\citeproc{ref-wilson2020case}{Wilson 2020})\footnote{For
  clarity: we follow standard ML convention in using ``degrees of
  freedom'' to refer to the number of parameters estimated from data.}.
Based purely on predictive performance, these solutions may seem to
provide compelling explanations for the data, when in fact they are
based on purely associative, semantically meaningless patterns. This
poses two related challenges: firstly, it makes these models inherently
opaque, since humans cannot simply interpret what type of explanation
the complex learned representations correspond to; secondly, even if we
could resolve the first challenge, it is not obvious how to mitigate
models from learning representations that correspond to meaningless and
implausible explanations.

@@

To our surprise, the second challenge has not yet attracted any
consolidated research effort. Specifically, there has been no concerted
effort towards improving model \textbf{explainability}, which we define
here as the degree to which learned representations correspond to
explanations that are interpretable and deemed \textbf{plausible} by
humans (see Definition~\ref{def-explainability}). Instead, the choice
has typically been to improve the capacity of XAI tools to identify the
subset explanations that are both plausible and valid for any given
model, independent of whether the learned representations are also
compatible with implausible explanations
(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
Fortunately, recent findings indicate that explainability can arise as
byproduct of regularization techniques aimed at other objectives such as
robustness, generalization and generative capacity Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}).

Building on these findings, we introduce \textbf{counterfactual
training}: a novel regularization technique geared explicitly towards
aligning model representations with plausible explanations. Our
contributions are as follows:

@@

To the best of our knowledge, our proposed framework for counterfactual
training represents the first attempt to use counterfactual explanations
during training to improve model explainability. In high-level terms, we
define model explainability as the extent to which valid explanations
derived for an opaque model are also deemed plausible with respect to
the underlying data and stakeholder requirements. To make this more
concrete, we follow Augustin, Meinke, and Hein
(\citeproc{ref-augustin2020adversarial}{2020}) in tieing the concept of
explainability to the quality of counterfactual explanations that we can
generate for a given model. The authors show that counterfactual
explanations---understood here as minimal input perturbations that yield
some desired model prediction---are generally more meaningful if the
underlying model is more robust to adversarial examples. We can make
intuitive sense of this finding when looking at adversarial training
(AT) through the lens of representation learning with high degrees of
freedom: by inducing models to ``unlearn'' representations that are
susceptible to worst-case counterfactuals (i.e.~adversarial examples),
AT effectively removes some implausible explanations from the solution
space.

@@

Improving the adversarial robustness of models is not the only path
towards aligning representations with plausible explanations. In a work
closely related to this one, Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}) show that explainability can
be improved through model averaging and refined model objectives. The
authors propose a way to generate counterfactuals that are maximally
\textbf{faithful} to the model in that they are consistent with what the
model has learned about the underlying data. Formally, they rely on
tools from energy-based modelling to minimize the divergence between the
distribution of counterfactuals and the conditional posterior over
inputs learned by the model. Their proposed counterfactual explainer,
\emph{ECCCo}, yields plausible explanations if and only if the
underlying model has learned representations that align with them. They
find that both deep ensembles
(\citeproc{ref-lakshminarayanan2016simple}{Lakshminarayanan, Pritzel,
and Blundell 2017}) and joint energy-based models (JEMs)
(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) tend to do
well in this regard.

Once again it helps to look at these findings through the lens of
representation learning with high degrees of freedom. Deep ensembles are
approximate Bayesian model averages, which are most called for when
models are underspecified by the available data
(\citeproc{ref-wilson2020case}{Wilson 2020}). Averaging across solutions
mitigates the aforementioned risk of relying on a single locally optimal
representations that corresponds to semantically meaningless
explanations for the data. Previous work by Schut et al.
(\citeproc{ref-schut2021generating}{2021}) similarly found that
generating plausible (``interpretable'') counterfactual explanations is
almost trivial for deep ensembles that have also undergone adversarial
training. The case for JEMs is even clearer: they involve a hybrid
objective that induces both high predictive performance and generative
capacity (\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}). This
is closely related to the idea of aligning models with plausible
explanations and has inspired our proposed counterfactual training
objective, as we explain in Section~\ref{sec-method}.

@@

Counterfactual training combines ideas from adversarial training,
energy-based modelling and counterfactuals explanations with the
explicit objective of aligning representations with plausible
explanations that comply with user requirements. In the context of CE,
plausibility has broadly been defined as the degree to which
counterfactuals comply with the underlying data generating process
(\citeproc{ref-poyiadzi2020face}{Poyiadzi et al. 2020};
\citeproc{ref-guidotti2022counterfactual}{Guidotti 2022};
\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). Plausibility
is a necessary but insufficient condition for using CE to provide
algorithmic recourse (AR) to individuals affected by opaque models in
practice. This is because for recourse recommendations to be
\textbf{actionable}, they need to not only result in plausible
counterfactuals but also be attainable. A plausible CE for a rejected
20-year-old loan applicant, for example, might reveal that their
application would have been accepted, if only they were 20 years older.
Ignoring all other features, this complies with the definition of
plausibility if 40-year-old individuals are in fact more credit-worthy
on average than young adults. But of course this CE does not qualify for
providing actionable recourse to the applicant. For our intents and
purposes, counterfactual training aims at improving model explainability
by aligning models with counterfactuals that meet both desiderata,
plausibility and actionability. Formally, we define explainability as
follows:

@@

Let \(M_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denote a supervised
classification model that maps from intuts to representations
\(\phi(x;theta)\) and finally to outputs. Assume that for any given
input-output pair \(\{\mathbf{x},\mathbf{y}\}\) there exists a
counterfactual
\(\mathbf{x}^{\prime} = \mathbf{x} + \Delta: M_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = M_\theta(\mathbf{x})\)
where \(\mathbf{y}^{+}\) denotes some target output. We say that
\(M_\theta\) is \textbf{explainable} to the extent that faithfully
generated counterfactuals are:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Plausible: \(\mathbf{x}^{\prime} \sim_{p} \mathcal{X}|\mathbf{y}^{+}\)
  with \(p \rightarrow 1\) as defined in Altmeyer et al.
  (\citeproc{ref-altmeyer2024faithful}{2024}).
\item
  Actionable: permutations \(\Delta\) are subject to actionability
  constraints.
\end{enumerate}

@@

\bibitem[\citeproctext]{ref-guidotti2022counterfactual}
Guidotti, Riccardo. 2022. {``Counterfactual Explanations and How to Find
Them: Literature Review and Benchmarking.''} \emph{Data Mining and
Knowledge Discovery}, 1--55.

@@

\bibitem[\citeproctext]{ref-poyiadzi2020face}
Poyiadzi, Rafael, Kacper Sokol, Raul Santos-Rodriguez, Tijl De Bie, and
Peter Flach. 2020. {``{FACE}: {Feasible} and Actionable Counterfactual
Explanations.''} In \emph{Proceedings of the {AAAI}/{ACM Conference} on
{AI}, {Ethics}, and {Society}}, 344--50.
